---
title: "Illustrative TV-HTE event study (with feedback decomposition)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Illustrative TV-HTE event study (with feedback decomposition)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5,
  warning   = FALSE,
  message   = FALSE
)
```

This vignette uses a synthetic panel to show the `tvhte` workflow. The panel has
persistent outcomes, heterogeneous dynamic responses, staggered adoption, and a
covariate that responds to past outcomes. We fit the structural TV-HTE model
from Botosaru and Liu (2025), then the homogeneous-feedback model from Botosaru
and Liu (2026), and use the two fitted objects to separate direct and indirect
dynamic responses.

The DGP here is synthetic only because the BL 2025 county-unemployment
analysis (their Section 6) does not ship replication data. The
workflow on real panel data is identical: swap the simulated `Y`,
`Y0`, `X`, `X0`, `t0` for your own panel arranged the same way.

## Setup and DGP

```{r dgp}
library(tvhte)
set.seed(11)

# Mixed cohort: half treated at t = 3, a quarter at t = 5, a quarter never.
N <- 1500; T <- 7
cohorts <- sample(c(3, 5, Inf), N, replace = TRUE, prob = c(0.5, 0.25, 0.25))

sim <- simulate_tvhte(
  N = N, T = T, t0 = cohorts, J = 3,
  rho_Y     = 0.45,   # persistent outcome
  rho_delta = 0.55,   # treatment effect decays with event time
  sigma_U   = 1, sigma_eps = 0.3,
  mu_alpha  = 0,      mu_delta0 = 1.5,    # average DATT at impact = 1.5
  sigma_alpha = 0.6,  sigma_delta0 = 0.5, # heterogeneity in both
  cor_alpha_delta = 0.3,
  beta = 0.4,
  feedback_gamma = c(intercept = 0.1, gamma_Y = 0.25,
                     gamma_X = 0.4,  sigma_eta = 0.3),
  seed = 11
)
str(sim, max.level = 1)
```

The simulated panel has `N = 1500` units observed at `t = 0, 1, ...,
7`. Three cohorts (treated at 3, treated at 5, never treated) are
mixed in known proportions. The covariate `X_{it}` adjusts each period
in response to lagged `Y` and lagged `X` — the homogeneous-feedback
DGP of BL 2026.

## Fit the structural model

```{r fit}
fit <- tvhte(
  Y  = sim$Y,
  Y0 = sim$Y0,
  t0 = sim$t0,
  J  = sim$J,
  X  = sim$X,
  compute_se = TRUE
)
print(fit)
```

The QMLE recovers the dynamic parameters cleanly: `rho_Y`, `rho_delta`,
the covariate coefficient `beta`, and the prior moments of the
correlated random coefficients `(alpha_i, delta_{i0})`. The "mean
posterior event-time effects" line shows the average across units of
the unit-level empirical-Bayes trajectories — this is what an event-
study plot would show *aggregated* across units.

## Unit-level posterior trajectories

The empirical-Bayes step gives a per-unit `(alpha_hat, delta0_hat)`
plus the full event-time path `delta_{i,j}` for `j = 0..J`.

```{r posterior, fig.height = 4}
library(graphics)
# Show 30 randomly-chosen treated units' posterior trajectories
treated_idx <- which(is.finite(sim$t0))
sub <- sample(treated_idx, 30)
matplot(0:fit$J, t(fit$delta_path[sub, ]),
        type = "l", col = adjustcolor("grey40", 0.4), lty = 1,
        xlab = "Event time j",
        ylab = "Posterior mean delta_{i,j}",
        main = "Unit-level event-time trajectories (30 random treated units)")
# Overlay the cross-unit mean
lines(0:fit$J, colMeans(fit$delta_path[treated_idx, ]),
      lwd = 2, col = "firebrick")
legend("topright", legend = c("unit-level posteriors", "average"),
       col = c("grey40", "firebrick"), lty = 1, lwd = c(1, 2), bty = "n")
```

The trajectories fan out because the random coefficient `delta_{i0}`
varies across units. The average path (red) decays at roughly the
estimated `rho_delta`.

## Fit the feedback model

```{r feedback}
fb <- fit_feedback(sim$Y, sim$Y0, sim$X, sim$X0)
print(fb)
```

The covariate `X_{it}` is modelled as an AR(1) with one lag of `Y` —
the simplest "homogeneous feedback" specification. Coefficients should
sit near the true `c(0.1, 0.25, 0.4)`.

Under the BL 2026 factorisation, the structural piece (fit by
`tvhte()`) and the feedback piece (fit by `fit_feedback()`) are
estimated separately — neither needs a parametric specification of the
other.

## Counterfactual decomposition

`simulate_counterfactual()` implements BL 2026's Algorithm 1: it
draws latent types from the estimated prior, evolves event-time
effects via the AR(1), and propagates `(Y, X)` jointly using the
estimated feedback process. The output is a joint counterfactual path
`(Y_i^{T,*}, X_i^{T,*})`, separating the **direct** effect (through
`delta_{i,j}`) from the **indirect** effect (through `X_t * beta`).

```{r counterfactual}
# Counterfactual 1: same units, treatment shifted earlier (t0 = 2)
cf_early <- simulate_counterfactual(fit, fb, t0_star = 2,
                                    N_star = 500, seed = 99)
# Counterfactual 2: same units, never treated
cf_never <- simulate_counterfactual(fit, fb, t0_star = Inf,
                                    N_star = 500, seed = 99)

# Difference at each calendar period gives the total dynamic response.
total_effect <- colMeans(cf_early$Y) - colMeans(cf_never$Y)

# Direct-only counterfactual: same treatment but freeze X at its baseline
# value (no feedback adjustment). We re-build by setting feedback gammas
# that ignore Y_lag and X_lag (so X stays at its intercept + noise).
fb_frozen <- fb
fb_frozen$coef["gamma_Y"] <- 0
fb_frozen$coef["gamma_X"] <- 0
cf_early_direct <- simulate_counterfactual(fit, fb_frozen, t0_star = 2,
                                           N_star = 500, seed = 99)
cf_never_direct <- simulate_counterfactual(fit, fb_frozen, t0_star = Inf,
                                           N_star = 500, seed = 99)
direct_effect <- colMeans(cf_early_direct$Y) - colMeans(cf_never_direct$Y)
indirect_effect <- total_effect - direct_effect

decomp <- data.frame(
  t = 1:T,
  total    = round(total_effect,    3),
  direct   = round(direct_effect,   3),
  indirect = round(indirect_effect, 3)
)
print(decomp)
```

```{r decomp-plot}
matplot(decomp$t, decomp[, -1],
        type = "l", lty = c(1, 2, 3), col = c("black", "firebrick", "steelblue"),
        lwd = 2, xlab = "Calendar time", ylab = "Effect on Y",
        main = "Dynamic decomposition: total vs direct + indirect")
abline(h = 0, lty = 3, col = "grey50")
legend("topleft", legend = c("Total", "Direct (delta)", "Indirect (X feedback)"),
       lty = c(1, 2, 3), col = c("black", "firebrick", "steelblue"),
       lwd = 2, bty = "n")
```

The total dynamic response (black) is the sum of the direct structural
effect (red dashed — what the unit would experience holding the
covariate path fixed) and the indirect adjustment effect (blue
dotted — how much of the response is mediated by the covariate
responding to past outcomes and treatment). In the BL 2026
minimum-wage example, the direct effect is the wage response holding
firm input demand fixed; the indirect effect is the additional
response that flows through firms re-optimising hours and employment
composition.

## On real data

For an applied dataset, the workflow is the same:

1. Arrange your panel as `(N x T)` outcome matrix `Y`, baseline vector
   `Y0`, optional `(N x T x K)` covariate array `X`, baseline `X0`,
   and length-`N` treatment-cohort vector `t0` (use `Inf` for
   never-treated units).
2. Call `tvhte(Y, Y0, t0, J, X)` for the structural fit.
3. Call `fit_feedback(Y, Y0, X, X0)` for the feedback fit.
4. Counterfactual decomposition with `simulate_counterfactual()`.

The QMLE step under the Gaussian working prior is consistent for the
common parameters even when the true prior on `(alpha_i, delta_{i0})`
is non-Gaussian (Botosaru and Liu 2025); the empirical-Bayes step
achieves ratio optimality. The homogeneous-feedback factorisation
delivers separately-identified direct and indirect effects (Botosaru
and Liu 2026).

## References

- Botosaru, Irene and Laura Liu (2025). "Time-Varying Heterogeneous
  Treatment Effects in Event Studies." [arXiv:2509.13698](https://arxiv.org/abs/2509.13698).
- Botosaru, Irene and Laura Liu (2026). "Event Studies with Feedback."
  *AEA Papers and Proceedings* 116: 70–74.
